---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nisone

<!-- badges: start -->
[![R-CMD-check](https://github.com/dcgerard/nisone/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dcgerard/nisone/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/dcgerard/nisone/graph/badge.svg?token=1PQ0SVRKKR)](https://codecov.io/gh/dcgerard/nisone)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

Provides different interval estimates of a location parameter when the sample size is one or more. These include classical methods when n=1, new Bayesian analogues of these classical methods, and extensions of these methods to larger sample sizes in the normal case. Other functions calculate Bayes factors based on t-statistics, and implement the (generalized) inverse normal distribution (and other inverse distributions). See Gerard (2026) for details of these methods.

## Installation

You can install the development version of nisone from [GitHub](https://github.com/dcgerard/nisone) with:

``` r
# install.packages("pak")
pak::pak("dcgerard/nisone", build_vignettes = TRUE)
```

# n=1 confidence interval

If we observe $X = 10$ and we have prior guess that the mean is around 5, then the resulting 95% CI based on a normal model is
```{r example}
library(nisone)
ci1(x = 10, A = 5)
```

Its full posterior distribution based on a probability matching prior can be calculated using the `n1post` family of functions. For example, the median and 95% credible interval is
```{r}
qn1post(p = c(0.025, 0.5, 0.975), A = 5, obs = 10, nu = 1, fam = "normal")
```
Some random posterior draws and density curve is
```{r}
samp <- rn1post(n = 10000, A = 5, obs = 10, nu = 1, fam = "normal")
samp <- samp[samp > -20 & samp < 40]
x <- seq(-20, 40, length.out = 500)
y <- dn1post(x = x, A = 5, obs = 10, nu = 1, fam = "normal")
graphics::hist(
  samp, 
  freq = FALSE, 
  breaks = 200, 
  border = "grey", 
  col = "grey",
  main = "Posterior Density and Histogram",
  xlab = "Mean",
  ylab = "Posterior Density"
)
graphics::lines(x, y, type = "l", col = "#E69F00")
```



If we just want to assume that $X$ comes from *any* symmetric unimodal distribution, a 95% CI for the mode is
```{r}
ci1(x = 10, A = 5, family = "uniform")
```

See more functionality by running
```r
browseVignettes("nisone")
```

# References

- Gerard, D. (2026). Constructing and extending *n* = 1 Bayesian confidence intervals of the normal mean. *In preparation*.
