---
title: "Introduction to nisone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nisone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
set.seed(1)
library(nisone)
```

This document introduces the main functions in the `{nisone}` R package. The most important of which are:

- `ci1()`: Frequentist $n=1$ confidence intervals.
- `wc_width()`: The half-width (in units of $|X-A|$) for the $n=1$ confidence intervals.
- `aug_t()`: Augmented $t$-interval, the generalization of `ci1()` for $n\geq 2$.
- `bci1()`: Bayesian version of `ci1()`.
- `blevel()`: The true confidence level of the credible interval from `bci1()`.
- `bcin()`: Bayesian interval for $n\geq 2$ assuming the coefficient of variation is exactly 1.
- `bft()`: Bayes factor given a $t$-statistic.
- `dn1post()`, `pn1post()`, `qn1post()`, `rn1post()`, `nun1post()`: The posterior distribution of the location parameter using the $n=1$ priors.
- `dinvnorm()`, `pinvnorm()`, `qinvnorm()`, `rinvnorm()`: The inverse normal distribution.
- `dginvnorm()`, `pginvnorm()`, `qginvnorm()`, `rginvnorm()`: The generalized inverse normal distribution.

These methods are all described in detail in Gerard (2026).

# $n=1$ Confidence Intervals

Suppose we have a single observations from a symmetric location-scale family. That is, $\rho()$ is any symmetric density function and the PDF of $X$ is $\frac{1}{\sigma}\rho\left(\frac{X-\mu}{\sigma}\right)$ for some center $\mu$ and some scale $\sigma$. For example, $X \sim N(\mu, \sigma^2)$. Given a prespecified value $A$, valid confidence intervals of the form
$$
X \pm \eta |X - A|
$$
and
$$
\frac{X + A}{2} \pm \eta |X - A|
$$
can be constructed to produce valid $(1 - \alpha)100\%$ confidence intervals. This is done by making $\eta$ large enough to bound the coverage probability below by $(1 - \alpha)$. See Blachman and Machol (1987).

The function `ci1()` produces these confidence intervals where $\rho()$ is (i) the standard normal function, (ii) Cauchy, and (iii) uniform. Surprisingly, constructing intervals using a uniform distribution makes them valid $(1 - \alpha)100\%$ confidence intervals when $X$ has *any* unimodal density.

Suppose we observe $X = 2$ and we use $A = 1$, then, centered at $X$, the various 95\% intervals are:
```{r}
ci1(x = 2, A = 1, type = "x", family = "cauchy")
ci1(x = 2, A = 1, type = "x", family = "normal")
ci1(x = 2, A = 1, type = "x", family = "uniform")
```
Centered at $(X + A) / 2$, they are
```{r}
ci1(x = 2, A = 1, type = "ave", family = "cauchy")
ci1(x = 2, A = 1, type = "ave", family = "normal")
ci1(x = 2, A = 1, type = "ave", family = "uniform")
```

If you are interested, you can get the values of $\eta$ for the normal, Cauchy, and uniform via `wc_width()`. E.g., for a normal distribution 95\% confidence interval of the form $(X + A)/2 \pm \eta|X-A|$, the value of $\eta$ is
```{r}
wc_width(alpha = 0.05, center = "ave", family = "normal")
```


# $n=1$ Bayesian Intervals

In Gerard (2026), we showed that (for $n=1$) using the following prior produces posterior credible intervals that are asymptotically valid confidence intervals.
$$
\pi(\mu,\nu) = |\mu - A|^{-1}\delta_{\nu^*}(\nu), \text{ where}\\
\nu = (\mu - A) / \sigma,\\
\nu^* = \mathrm{argmax}_{\nu>0}\nu\rho\left(\nu\right),
$$
and $\delta_{\nu^*}(\nu)$ is a pointmass at $\nu^*$. This is asymptotic *not* in the sample size, but in the confidence level. So the credible intervals are approximate confidence intervals for small $\alpha$, which is the typical case.

In the normal case, this corresponds to the prior
$$
\pi(\mu) = |\mu - A|^{-1}\\
\sigma^2 = (\mu - A)^2 \text{ w.p. 1}.
$$
The resulting posterior distribution is inverse normal (see below). We have implemented this Bayesian approach in `bci1()`. Though, the asymptotics are so good that you end up getting almost the exact same interval as the frequentist $n=1$ interval,

```{r}
bci1(x = 1, A = 0)
ci1(x = 1, A = 0)
```

The true level of the Bayesian credible intervals can be found by `blevel()`. It's almost exactly the correct level for 95\% confidence intervals. But they can dip a little further below the nominal level for smaller levels:
```{r}
blevel(level = 0.7, nu = 0.8)
```

# $n \geq 1$ Augmented t-intervals

Suppose we have data $X_1,\ldots,X_n \sim N(\mu, \sigma^2)$. Let $A$ be some prespecified value. Let $\hat{\mu}$ and $\hat{\sigma}^2$ be the
sample mean and sample variance using the augmented data (all the $X$'s and $A$). We consider intervals of the form
$$
\hat{\mu} \pm \eta \hat{\sigma}/\sqrt{n + 1},
$$
where $\eta$ is chosen to control the error probability. We call these "augmented $t$-interval". The function that calculates them is `aug_t()`.

If $A$ is close to $\mu$, the augmented $t$-intervals tend to be smaller on average. For $n=2$, you only need $A$ to be within about 4 standard deviations of $\mu$ to see large improvements.
```{r}
n <- 2
mu <- 10
sigma <- 4
A <- 5
x <- rnorm(n = n, mean = mu, sd = sigma)
aug_t(x = x, A = A)
t.test(x)$conf.int
```


An augmented $t$-interval can be viewed as an inverted frequentist test that uses a Bayes factor as a frequentest test statistic. Gronau and Wagenmakers (2020) studied the connection between Bayes factors and *t*-statistics. Let $\delta$ be the standardized effect size, let $\sigma^2$ be the variance (assumed equal in two-sample case). In the one-sample case, $\delta = \frac{\mu - \mu_0}{\sigma}$, where $\mu_0$ is the null value. In the two-sample case, $\delta = \frac{\mu_1 - \mu_2}{\sigma}$, where $\mu_1$ and $\mu_2$ are the means of the two-samples. We place the prior $1/\sigma^2$ under the null and $\pi(\delta)/\sigma^2$ under the alternative, for some arbitrary density $\pi(\cdot)$. Given this setting, the Bayes factor is a function of the $t$-statistic. 
$$
\mathrm{BF} = \frac{\int_{\delta}T_{\nu}(t|\sqrt{n_\delta}\delta)\pi(\delta)\mathrm{d}\delta}{T_{\nu}(t)},
$$
where $T_{\nu}()$ is the central $t$-density with $\nu$ degrees of freedom, and $T_{\nu}(|a)$ is the non-central $t$-density with $\nu$ degrees of freedom and non-centrality parameter $a$. In the one-sample case, $n_{\delta} = n$ and $\nu = n - 1$, and in the two-sample case $n_{\delta} = \left(\frac{1}{n_1} + \frac{1}{n_2}\right)^{-1}$ and $\nu = n_1 + n_2 - 2$. The `bft()` function will calculate this Bayes factor given a $t$-statistic and any prior distribution (absolutely continuous with respect to Lebesgue measure) you provide over $\delta$ (with Cauchy as a default). 
```{r}
tstat <- mean(x) / (sd(x) / length(x))
bft(t = tstat, nu = length(x) - 1, nd = length(x))
```
To get the Bayes factor using augmented data, you assume that $\mu$ has a normal prior with mean $A$ and variance $\sigma^2$, then you just include $A$ in calculating the augmented $t$-statistic
```{r}
x_aug <- c(x, A)
tstat <- mean(x_aug) / (sd(x_aug) / length(x_aug))
bft(t = tstat, nu = length(x_aug) - 1, nd = length(x_aug))
```

# Generalized Inverse Normal Distribution

Robert (1991) considered the generalized inverse normal distribution of the form
$$
f(x|\alpha,\mu,\tau) = K(\alpha,\mu,\tau)|x|^{-\alpha}\exp\left\{-\frac{1}{2\tau^2}\left(\frac{1}{x}-\mu\right)^2\right\},
$$
where $K(\alpha,\mu,\tau)$ is the proportionality constant. When $\alpha = 2$, this corresponds to the inverse normal distribution (different from the inverse Gaussian distribution), where $\frac{1}{X} \sim N(\mu, \tau^2)$. The generalized inverse normal distribution shows up in Bayesian analysis when you parameterize the normal distribution in terms of its mean $\mu$ and its coefficient of variation $\nu = \mu / \tau$.

The `nisone()` R package contains density, distribution, quantile, and random generation functions for the (generalized) inverse normal distribution. You can use this to get the full posterior distribution for $n=1$ intervals when $\rho()$ is normal. In the normal case, the marginal posterior of $\mu - A$ is inverse normal with inverse mean $\frac{1}{X - A}$ and inverse variance $\frac{1}{(X - A)^2}$. E.g., if we observed $X = 2$ and use $A = 0.75$, the full posterior density is
```{r}
museq <- seq(-2, 5, length.out = 500)
X <- 2
A <- 0.75
density <- dinvnorm(x = museq - A, imean = 1 / (X - A), isd = 1 / abs(X - A))
graphics::plot(
  museq, 
  density,
  type = "l",
  xlab = expression(mu),
  ylab = "Posterior Density"
)
```

You can get 95\% Credible intervals via
```{r}
qinvnorm(p = c(0.025, 0.975), imean = 1 / (X - A), isd = 1 / abs(X - A)) + A
```
Which agree with `bci1()`
```{r}
bci1(x = X, A = A)
```

Interfacing with the posterior is provided by the `n1post` functions so that you don't need to manually calculate the posterior above.
```{r}
dn1post(x = 10, obs = X, A = A, nu = 1, fam = "normal")
dinvnorm(x = 10 - A, imean = 1 / (X - A), isd = 1 / abs(X - A))

qn1post(p = c(0.025, 0.975), A = A, obs = X, nu = 1, fam = "normal")
qinvnorm(p = c(0.025, 0.975), imean = 1 / (X - A), isd = 1 / abs(X - A)) + A
```

More generally, we have functions for the generalized inverse normal.
```{r}
#| fig-width: 5
#| fig.height: 3
set.seed(50)
qginvnorm(p = 0.025, alpha = 4, mu = 0.5, tau = 1)
pginvnorm(q = -1.345, alpha = 4, mu = 0.5, tau = 1)
samp <- rginvnorm(n = 1000, alpha = 4, mu = 0.5, tau = 1)
x <- seq(min(samp), max(samp), length.out = 500)
y <- dginvnorm(x = x, alpha = 4, mu = 0.5, tau = 1)
modes <- xginvnorm(alpha = 4, mu = 0.5, tau = 1)
hist(
  samp,
  freq = FALSE,
  breaks = 100,
  xlab = "x",
  main = "Generalized Inverse Normal Density")
lines(x, y, col = "#E69F00")
abline(v = modes, col = "#56B4E9", lty = 2)
```

# References

- Blachman, N., & Machol, R. (1987). Confidence intervals based on one or more observations. *IEEE Transactions on Information Theory*, 33(3), 373-382. [doi:10.1109/TIT.1987.1057306](https://doi.org/10.1109/TIT.1987.1057306)
- Gerard, D. (2026). Constructing and extending *n* = 1 Bayesian confidence intervals for location parameters in location-scale families. *In preparation*.
- Gronau, Q. F., Ly, A., & Wagenmakers, E. J. (2020). Informed Bayesian *t*-Tests. *The American Statistician*, 74(2), 137–143. [doi:10.1080/00031305.2018.1562983](https://doi.org/10.1080/00031305.2018.1562983)
- Robert, C. (1991). Generalized inverse normal distributions. *Statistics & Probability Letters*, 11(1), 37-41. [doi:10.1016/0167-7152(91)90174-P](https://doi.org/10.1016/0167-7152(91)90174-P)
