---
title: "Introduction to nisone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to nisone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nisone)
```

This document introduces the main functionality in the `{nisone}` R package. The most important of which are:

- `ci1()`: Frequentist $n=1$ confidence interval
- `aug_t()`: Augmented $t$-interval, the generalization of `ci1()` for $n\geq 2$.
- `bci1()`: Bayesian version of `ci1()`
- `blevel()`: The true confidence level of the credible interval from `bci1()`
- `bcin()`: Bayesian interval for $n\geq 2$ assuming the coefficient of variation is exactly 1.
- `bft()`: Bayes factor given a $t$-statistic
- `dinvnorm()`, `pinvnorm()`, `qinvnorm()`, `rinvnorm()`: The inverse normal distribution.
- `dginvnorm()`, `pginvnorm()`, `qginvnorm()`, `rginvnorm()`: The generalized inverse normal distribution.

# $n=1$ Confidence Intervals

Suppose we have a single observations from a symmetric location-scale family. That is, $\rho()$ is any symmetric density function and the PDF of $X$ is $\frac{1}{\sigma}\rho\left(\frac{X-\mu}{\sigma}\right)$ for some center $\mu$ and some scale $\sigma$. For example, $X \sim N(\mu, \sigma^2)$. Given a prespecified value $A$, valid confidence intervals of the form
$$
X \pm \eta |X - A|
$$
and
$$
\frac{X + A}{2} \pm \eta |X - A|
$$
can be constructed to produce valid $(1 - \alpha)100\%$ confidence intervals. This is done by making $\eta$ large enough to bound the coverage probability below by $(1 - \alpha)$. See Blachman and Machol (1987).

The function `ci1()` produces these confidence intervals where $\rho()$ is (i) the standard normal function, (ii) Cauchy, and (iii) uniform. Surprisingly, constructing intervals using a uniform distribution makes them valid $(1 - \alpha)100\%$ confidence intervals when $X$ has *any* unimodal density.

Suppose we observe $X = 2$ and we use $A = 1$, then, centered at $X$, the various 95\% intervals are:
```{r}
ci1(x = 2, A = 1, type = "x", family = "cauchy")
ci1(x = 2, A = 1, type = "x", family = "normal")
ci1(x = 2, A = 1, type = "x", family = "uniform")
```
Centered at $(X + A) / 2$, they are
```{r}
ci1(x = 2, A = 1, type = "ave", family = "cauchy")
ci1(x = 2, A = 1, type = "ave", family = "normal")
ci1(x = 2, A = 1, type = "ave", family = "uniform")
```

# $n=1$ Bayesian Intervals

In Gerard (2026), we showed that using the following prior produces posterior credible intervals that are asymptotically valid confidence intervals.
$$
\pi(\mu,\nu) = |\mu - A|^{-1}\delta_{\nu^*}(\nu), \text{ where}\\
\nu = (\mu - A) / \sigma,\\
\nu^* = \mathrm{argmax}_{\nu>0}\nu\rho\left(\nu\right),
$$
and $\delta_{\nu^*}(\nu)$ is a pointmass at $\nu^*$. This is asymptotic *not* in the sample size, but in the confidence level. So the credible intervals are approximate confidence intervals for small $\alpha$, which is the typical case.

In the normal case, this corresponds to the prior
$$
\pi(\mu) = |\mu - A|^{-1}\\
\sigma^2 = (\mu - A)^2 \text{ w.p. 1}.
$$
The resulting posterior distribution is inverse normal (see below). We have implemented this Bayesian approach in `bci1()`. Though, the asymptotics are so good that you end up getting almost the exact same interval as the frequentist $n=1$ interval,

```{r}
bci1(x = 1, A = 0)
ci1(x = 1, A = 0)
```

The true level of the Bayesian credible intervals can be found by `blevel()`. It's almost exactly the correct level for 95\% confidence intervals. But they can dip a little further below the nominal level for smaller levels:
```{r}
blevel(level = 0.7, nu = 0.8)
```

# $n \geq 1$ Augmented t-intervals


See Gronau and Wagenmakers (2020) for more information on Bayes factors and *t*-statistics.

# Generalized Inverse Normal Distribution

Robert (1991) considered the generalized inverse normal distribution of the form
$$
f(x|\alpha,\mu,\tau) = K(\alpha,\mu,\tau)|x|^{-\alpha}\exp\left\{-\frac{1}{2\tau^2}\left(\frac{1}{x}-\mu\right)^2\right\},
$$
where $K(\alpha,\mu,\tau)$ is the proportionality constant. When $\alpha = 2$, this corresponds to the inverse normal distribution (different from the inverse Gaussian distribution), where $\frac{1}{X} \sim N(\mu, \tau^2)$. The generalized inverse normal distribution shows up in Bayesian analysis when you parameterize the normal distribution in terms of its mean $\mu$ and its coefficient of variation $\nu = \mu / \tau$.

The `nisone()` R package contains density, distribution, quantile, and random generation functions for the (generalized) inverse normal distribution. You can use this to get the full posterior distribution for $n=1$ intervals when $\rho()$ is normal. In the normal case, the marginal posterior of $\mu - A$ is inverse normal with inverse mean $\frac{1}{X - A}$ and inverse variance $\frac{1}{(X - A)^2}$. E.g., if we observed $X = 2$ and use $A = 1$, the full posterior density is
```{r}
museq <- seq(-2, 5, length.out = 500)
X <- 2
A <- 1
density <- dinvnorm(x = museq - A, imean = 1 / (X - A), isd = 1 / (X - A)^2)
plot(
  museq, 
  density,
  type = "l",
  xlab = expression(mu),
  ylab = "Posterior Density")
```

You can get 95\% Credible intervals via
```{r}
qinvnorm(p = c(0.025, 0.975), imean = 1 / (X - A), isd = 1 / (X - A)^2) + A
```
Which agree with `bci1()`
```{r}
bci1(x = 2, A = 1)
```

More generally, we have functions for the generalized inverse normal.
```{r}
#| fig-width: 5
#| fig.height: 3
set.seed(50)
qginvnorm(p = 0.025, alpha = 4, mu = 0.5, tau = 1)
pginvnorm(q = -1.345, alpha = 4, mu = 0.5, tau = 1)
samp <- rginvnorm(n = 1000, alpha = 4, mu = 0.5, tau = 1)
x <- seq(min(samp), max(samp), length.out = 500)
y <- dginvnorm(x = x, alpha = 4, mu = 0.5, tau = 1)
modes <- xginvnorm(alpha = 4, mu = 0.5, tau = 1)
hist(
  samp,
  freq = FALSE,
  breaks = 100,
  xlab = "x",
  main = "Generalized Inverse Normal Density")
lines(x, y, col = "#E69F00")
abline(v = modes, col = "#56B4E9", lty = 2)
```


# References

- Blachman, N., & Machol, R. (1987). Confidence intervals based on one or more observations. *IEEE Transactions on Information Theory*, 33(3), 373-382. [doi:10.1109/TIT.1987.1057306](https://doi.org/10.1109/TIT.1987.1057306)
- Gerard, D. (2026). Constructing and extending *n* = 1 Bayesian confidence intervals of the normal mean. *In preparation*.
- Gronau, Q. F., Ly, A., & Wagenmakers, E. J. (2020). Informed Bayesian *t*-Tests. *The American Statistician*, 74(2), 137–143. [doi:10.1080/00031305.2018.1562983](https://doi.org/10.1080/00031305.2018.1562983)
- Robert, C. (1991). Generalized inverse normal distributions. *Statistics & Probability Letters*, 11(1), 37-41. [doi:10.1016/0167-7152(91)90174-P](https://doi.org/10.1016/0167-7152(91)90174-P)
